Modal Abstractions for Virtualizing Memory Addresses

### 2023-03-02 programlogic.tex
+
sync with Desk

Therefore, an address-space, as an abstraction, can be treated as naming the memory state as a modal frame and the choice of page table root as a world in Kripke-style semantics. Not suprisingly, transition between two address-spaces, then, can be just an entailment relation \textit{alternating} the \textit{named-state} against the address-space invariant ($\mathcal{I}\textsf{ASpace}$ in Figure \ref{fig:peraspaceinvariant}). 
\[
  \begin{array}{l}
\hbox{(\TirNameStyle{ModalAddressSpace})} \qquad
 [r](P)  \;\stackrel{\triangle}{=} (\mathcal{W}_{64} -\textsf{b}> \textsf{uPredI} (\textsf{iResUR} \;\Sigma)) -\textsf{b}> \textsf{uPredI} \_ := \lambda \textsf{P},\; \textsf{P r}. 
  \end{array}
\]
Being inspired by what hybrid logic \ref{} calls a satisfaction operator, which evaluates the truth of a predicate in a named alternative state (here, address space), we give a modal definition describing the truth of assertions for the resources (i.e. virtual pointsto relations) inside the address space. Ignoring the predicate types for a while, $[r]P$ in Definition $\hbox{(\TirNameStyle{ModalAddressSpace})}$, indicates that $P$ holds in the virtual address space rooted at $r$, the truth ($P$) on an address-space is indexed by the root page-table address $r$ of the address-space. In the rest of this section, we explain the structural aspects -- the modal resource context of address space modality -- , and how we lift the interaction of address-space modality with the ambient logic, i.e. separation logic as an entailment for specifying \textit{address-space-switch}.  
\paragraph{Modal Resource Context}
\label{sec:resourcecontext}
When we pick up a certain contingency, which is \textit{"a certain virtual-pointsto facts whose validity is indexed by the root address of an address space"}for address space modality, we, in fact, make our choice around some logical resources for which we care about reasoning, which is virtual pointsto relation shown in Figure \ref{fig:virtualpointstosharing}. In other words, our modality definition is the logical virtual pointsto resources: introduction/elimination of logical virtual pointsto resources are the ones essential to the reasoning capabilities the address space modality provides, i.e. validity of virtual addresses under/out of modality -- resources under/out of \textit{modal resource context}.

The choice of resource defining the modal context already exhibits itself as a reasoning principle (Figure \ref{fig:modalcontextinvariance}) in which we can intro/eliminate any resource that is not virtual pointsto into/from modal context without dischargin any obligation. We can use this entailment for  \todo[inline,color=red]{Ask Colin}.
  \[
  \begin{array}{l}
\hbox{(\TirNameStyle{ModalContextInvariance})} \qquad
 [r](Q \star \ell \mapsto_{\textsf{p}} \vpage)  \;\vdash [r](Q) \star \ell \mapsto_{\textsf{p}} \vpage 
  \end{array}
  \]
\begin{remark}[The Choice of Modal Context Definition as a Pattern of Verification Context]
  \label{remark:pattern}
  By choosing the contingency and what defines the modal contex, we simply introduce 
  \begin{enumerate}
  \item the pattern of what the essential abstraction (e.g. address spaces) we care about to do verification for is
  \item how a modal context (e.g. a address-space modality $[r]P$), logically, draws the attention to the certain facts in the interest of reasoning (e.g. virtual pointsto relations in the current address-space in Figure \ref{fig:virtualpointstosharing})
  \item and, offloads the burden of individual bookkeeping of these facts (e.g. virtual pointsto relations per address-space) under different context (e.g. virtual pointsto relations in other address-spaces) via utilizing the \textit{summarization} aspect of the contingency it represents (e.g. other virtual pointsto relations are valid under other certain address-space page-table root addresseses).  
  \end{enumerate}
\end{remark}
\paragraph{Interaction}
\label{sec:interaction}
So far, we have just mentioned the structural properties of our modal definition (Remark \ref{remark:pattern}), and what inherent entailments (Rule \hbox{(\TirNameStyle{ModalContextInvariance})}) these properties provide. However, all these structural aspects of using the address-space modality ($[r]P$) considers the fixed index ($r$) of on the truth $P$. From the client's perspective, structural properties of the modal abstraction mainly constitutes the modal frame with summarizing the facts for one another address-space under the address-space modality roughly as it is in the following spec
\[
\{ \crthree \mapsto_{r} r0 \star \textsf{va1} \mapsto_{v} \textsf{v}_1 \star \textsf{va2} \mapsto_{v} \textsf{v}_2 \star \textsf{va3} \mapsto_{v} \textsf{v}_3 \star [r1]\;P
\]
where we see the \textit{current address-space} rooted at $r0$ is loaded into the  \textit{current-view} of the memory with its resources ($ \textsf{va1} \mapsto_{v} \textsf{v}_1 \star \textsf{va2} \mapsto_{v} \textsf{v}_2 \star \textsf{va3} \mapsto_{v} \textsf{v}_3$), and framed by the address-space rooted at $r1$ with resources $P$. 
Now, we need to a rule capturing the notion that switching the address spaces by pulling the resources ($P$) out of the one under modality into the \textit{“current view”} of memory, and leaves other assertions true relative to the previous address space (the one rooted at $r0$). To do so, we need to have a proof rule for the instruction (\lstinline|mov\_ctl|) updating the $\crthree$ register.

\subsection{Considering Hoare Doubles in the Context of Switching Address-Space}
An important subtlety arises with supporting \lstinline|mov|s into \lstinline|%cr3|. Consider the hypothetical rule:
\begin{mathpar}
\inferrule[Broken]{ }{
  \{P \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
  \{[r_1](P) \ast cr3=r_2 \ast r = r_2 \ast Q\}
}
\end{mathpar}
This rule captures the intuitive change of address space in a Hoare triple, rather than double, form. The problem with this is that it interacts quite poorly with the traditional frame rule and the modal flavor of virtual points-to assertions:
\begin{mathpar}
  \inferrule*[right=Frame]{
    \inferrule*[right=Broken]{ }{
    \{\mathsf{emp} \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{[r_1](\mathsf{emp}) \ast cr3=r_1 \ast r=r_2 \ast Q\}
    }
  }{
    \{a\mapsto_\mathsf{v} x \ast \mathsf{emp} \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{a\mapsto_\mathsf{v} x \ast [r_1](\mathsf{emp}) \ast cr3=r_1 \ast r=r_2 \ast Q\}
  }
\end{mathpar}
Notice that both the precondition and postcondition assert that $a\mapsto_\mathsf{v} x$ in the current address space, but we have no basis for concluding that address translation is preserved by the change of address space. So this derivation clearly leads to an unsound conclusion. This suggestss that the traditional frame rule and the Hoare triple presentation of the change-of-address-space rule cannot soundly coexist in the same system.
The heart of the problem is that while updating \lstinline|cr3| is \emph{physically} local, it globally changes the interpretation of virtual addresses. So it is simply unsound to frame around \lstinline|cr3| updates.

@@

This is then how we justify our actual rule for \lstinline|cr3| updates:
\begin{mathpar}
\inferrule[ChangeAddressSpace]{
  \{[r_1](P) \ast cr3=r_2 \ast r = r_2 \ast Q\}\overline{is}
}{
  \{P \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r;\;\overline{is}
  %\lstinline|mov %cr3, r| 
}
\end{mathpar}
Because the precondition on this rule is global, we avoid issues with framing.

@@

Readers familiar with dynamic frames~\cite{parkinson2011relationship} might find it useful to notice that a different perspective on this matter is that virtual points-to assertions are self-stable \emph{except} for changes in \lstinline|cr3|, so framing would then naturally require other means of holding \lstinline|cr3| constant (or saving and restoring it).
Virtual points-to assertions could be made self-stable by also giving them partial ownership over \lstinline|cr3| assertions, but this would require explicitly plumbing that ownership from \emph{all} assertions back to any place an address space change might occur; this would seem to be a far graver loss of modularity than this extra quirk in framing discussions.

\begin{figure}
\begin{mathpar}
\inferrule[ChangeAddressSpace]{
  \{[r_1](P) \ast cr3=r_2 \ast r = r_2 \ast Q\}\overline{is}
}{
  \{P \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r;\;\overline{is}
  %\lstinline|mov %cr3, r| 
}
\end{mathpar}
\caption{Selected Reasoning Rules for ..}
\label{fig:wpdamd}
\end{figure}

>>>>

Therefore, an address-space, as an abstraction, can be treated as naming the memory state as a modal frame and the choice of page table root as a world in Kripke-style semantics. Not suprisingly, transition between two address-spaces, then, can be just an entailment relation \textit{alternating} the \textit{named-state} against the address-space invariant ($\mathcal{I}\textsf{ASpace}$ in Figure \ref{fig:peraspaceinvariant}). 
\[
  \begin{array}{l}
    \hbox{(\TirNameStyle{ModalAddressSpace})} \\ \qquad
         [r](P)  \;\stackrel{\triangle}{=}  (\mathcal{W}_{64} -\textsf{b}> \textsf{uPredI} (\textsf{iResUR} \;\Sigma)) -\textsf{b}> \textsf{uPredI} \_ := \lambda \textsf{P},\; \textsf{P r}. 
  \end{array}
\]
Being inspired by what hybrid logic \ref{} calls a satisfaction operator, which evaluates the truth of a predicate in a named alternative state (here, address space), we give a modal definition describing the truth of assertions for the resources (i.e. virtual pointsto relations) inside the address space. Ignoring the predicate types for a while, $[r]P$ in Definition $\hbox{(\TirNameStyle{ModalAddressSpace})}$, indicates that $P$ holds in the virtual address space rooted at $r$, the truth ($P$) on an address-space is indexed by the root page-table address $r$ of the address-space. In the rest of this section, we explain the structural aspects -- the modal resource context of address space modality -- , and how we lift the interaction of address-space modality with the ambient logic, i.e. separation logic as an entailment for specifying \textit{address-space-switch}.  
\paragraph{Modal Resource Context}
\label{sec:resourcecontext}
When we pick up a certain contingency, which is \textit{"a certain virtual-pointsto facts whose validity is indexed by the root address of an address space"}for address space modality, we, in fact, make our choice around some logical resources for which we care about reasoning, which is virtual pointsto relation shown in Figure \ref{fig:virtualpointstosharing}. In other words, our modality definition is the logical virtual pointsto resources: introduction/elimination of logical virtual pointsto resources are the ones essential to the reasoning capabilities the address space modality provides, i.e. validity of virtual addresses under/out of modality -- resources under/out of \textit{modal resource context}.

The choice of resource defining the modal context already exhibits itself as a reasoning principle (Figure \ref{fig:modalcontextinvariance}) in which we can intro/eliminate any resource that is not virtual pointsto into/from modal context without dischargin any obligation. We can use this entailment for  \todo[inline,color=red]{Ask Colin}.
  \[
  \begin{array}{l}
\hbox{(\TirNameStyle{ModalContextInvariance})} \qquad
 [r](Q \star \ell \mapsto_{\textsf{p}} \vpage)  \;\vdash [r](Q) \star \ell \mapsto_{\textsf{p}} \vpage 
  \end{array}
  \]
\begin{remark}[The Choice of Modal Context Definition as a Pattern of Verification Context]
  \label{remark:pattern}
  By choosing the contingency and what defines the modal contex, we simply introduce 
  \begin{enumerate}
  \item the pattern of what the essential abstraction (e.g. address spaces) we care about to do verification for is
  \item how a modal context (e.g. a address-space modality $[r]P$), logically, draws the attention to the certain facts in the interest of reasoning (e.g. virtual pointsto relations in the current address-space in Figure \ref{fig:virtualpointstosharing})
  \item and, offloads the burden of individual bookkeeping of these facts (e.g. virtual pointsto relations per address-space) under different context (e.g. virtual pointsto relations in other address-spaces) via utilizing the \textit{summarization} aspect of the contingency it represents (e.g. other virtual pointsto relations are valid under other certain address-space page-table root addresseses).  
  \end{enumerate}
\end{remark}
\paragraph{Interaction}
\label{sec:interaction}
So far, we have just mentioned the structural properties of our modal definition (Remark \ref{remark:pattern}), and what inherent entailments (Rule \hbox{(\TirNameStyle{ModalContextInvariance})}) these properties provide. However, all these structural aspects of using the address-space modality ($[r]P$) considers the fixed index ($r$) of on the truth $P$. From the client's perspective, structural properties of the modal abstraction mainly constitutes the modal frame with summarizing the facts for one another address-space under the address-space modality roughly as it is in the following spec
\[
\{ \crthree \mapsto_{r} r0 \star \textsf{va1} \mapsto_{v} \textsf{v}_1 \star \textsf{va2} \mapsto_{v} \textsf{v}_2 \star \textsf{va3} \mapsto_{v} \textsf{v}_3 \star [r1]\;P \}
\]
where we see the \textit{current address-space} rooted at $r0$ is loaded into the  \textit{current-view} of the memory with its resources ($ \textsf{va1} \mapsto_{v} \textsf{v}_1 \star \textsf{va2} \mapsto_{v} \textsf{v}_2 \star \textsf{va3} \mapsto_{v} \textsf{v}_3$), and framed by the address-space rooted at $r1$ with resources $P$. 
Now, we need to a rule capturing the notion that switching the address spaces by pulling the resources ($P$) out of the one under modality into the \textit{“current view”} of memory, and leaves other assertions true relative to the previous address space (the one rooted at $r0$). To do so, we need to have a proof rule for the instruction (\lstinline|mov_ctl|) updating the $\crthree$.

\subsection{Considering Hoare Doubles in the Context of Switching Address-Space}
An important subtlety arises with supporting \lstinline|mov|s into \lstinline|%cr3|. Consider the hypothetical rule:
\begin{mathpar}
\inferrule[Broken]{ }{
  \{P \ast cr3\mapsto_{r} \;r_1 \ast r \mapsto_{r} \;r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
  \{[r_1](P) \ast cr3 \mapsto_{r} r_2 \ast r \mapsto_{r} r_2 \ast Q\}
}
\end{mathpar}
This rule captures the intuitive change of address space in a Hoare triple, rather than double, form. The problem with this is that it interacts quite poorly with the traditional frame rule and the modal flavor of virtual points-to assertions:
\begin{mathpar}
  \inferrule*[right=Frame]{
    \inferrule*[right=Broken]{ }{
    \{\mathsf{emp} \ast cr3 \mapsto_{r} r_1 \ast r\mapsto_{r} r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{[r_1](\mathsf{emp}) \ast cr3 \mapsto_{r} r_1 \ast r \mapsto_{r} r_2 \ast Q\}
    }
  }{
    \{a\mapsto_\mathsf{v} x \ast \mathsf{emp} \ast cr3\mapsto_{r}r_1 \ast r \mapsto_{r}r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{a\mapsto_\mathsf{v} x \ast [r_1](\mathsf{emp}) \ast cr3 \mapsto_{r} r_1 \ast r \mapsto_{r} r_2 \ast Q\}
  }
\end{mathpar}
Notice that both the precondition and postcondition assert that $a\mapsto_\mathsf{v} x$ in the current address space, but we have no basis for concluding that address translation is preserved by the change of address space. So this derivation clearly leads to an unsound conclusion. This suggestss that the traditional frame rule and the Hoare triple presentation of the change-of-address-space rule cannot soundly coexist in the same system.
The heart of the problem is that while updating \lstinline|cr3| is \emph{physically} local, it globally changes the interpretation of virtual addresses. So it is simply unsound to frame around \lstinline|cr3| updates.

@@

This is then how we justify our actual rule for \lstinline|cr3| updates:
\begin{mathpar}
\inferrule[ChangeAddressSpace]{
  \{[r_1](P) \ast cr3 \mapsto_{r} r_2 \ast r \mapsto_{r} r_2 \ast Q\}\overline{is}
}{
  \{P \ast cr3 \mapsto_{r} r_1 \ast r \mapsto_{r} r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r;\;\overline{is}
  %\lstinline|mov %cr3, r| 
}
\end{mathpar}
Because the precondition on this rule is global, we avoid issues with framing.

@@

Readers familiar with dynamic frames~\cite{parkinson2011relationship} might find it useful to notice that a different perspective on this matter is that virtual points-to assertions are self-stable \emph{except} for changes in \lstinline|cr3|, so framing would then naturally require other means of holding \lstinline|cr3| constant (or saving and restoring it).
Virtual points-to assertions could be made self-stable by also giving them partial ownership over \lstinline|cr3| assertions, but this would require explicitly plumbing that ownership from \emph{all} assertions back to any place an address space change might occur; this would seem to be a far graver loss of modularity than this extra quirk in framing discussions.
\todo{inlined, color= red}[STOP HERE, go back fix the stuff discussed with COLIN]
\begin{figure}
\begin{mathpar}
\inferrule[WriteToMemoryFromReg]{
  \{[r_1](P) \ast cr3=r_2 \ast r = r_2 \ast Q\}\overline{is}
}{
  \{P \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r;\;\overline{is}
  %\lstinline|mov %cr3, r| 
} 
\inferrule[WriteToRegFromReg]{
  \{[r_1](P) \ast cr3=r_2 \ast r = r_2 \ast Q\}\overline{is}
}{
  \{P \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r;\;\overline{is}
  %\lstinline|mov %cr3, r| 
} \\ 
\inferrule[WriteToRegFromMem]{
  \{[r_1](P) \ast cr3=r_2 \ast r = r_2 \ast Q\}\overline{is}
}{
  \{P \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r;\;\overline{is}
  %\lstinline|mov %cr3, r| 
}
\inferrule[WriteToCtlFromReg]{
  \{[r_1](P) \ast cr3=r_2 \ast r = r_2 \ast Q\}\overline{is}
}{
  \{P \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r;\;\overline{is}
  %\lstinline|mov %cr3, r| 
}
\end{mathpar}
\caption{Selected Reasoning Rules for ..}
\label{fig:wpdamd}
\end{figure}
